Everything you wanted to know about PanDA but were truly truly afraid to ask

### 2017-03-07 PandaEvolution.tex
Copy and address Sergey's comments + bibliography

PanDA WMS has been designed to support distributed computing. Executing ATLAS
workloads or workflows involves concurrent and/or sequential runs of possibly
large amount of jobs, each requiring no or minimal parallelization and no
runtime communication. Thus, computing infrastructure like WLCG have been
designed to aggregate large amount of computing resources across sites, often
geographically dislocated. While each site may deploy runtime message-passing
capabilities, usually these are not used to perform distributed computations.

@@

Currently, no HEP application can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks can be offloaded from
the Grid to Titan's large amount of cores. Further, when HEP tasks can be
partitioned to independent jobs, Titan work nodes can be used to execute up to
16 concurrent jobs' payload, one for each available core. Given these
constraints and challenges, the type of task most suitable for execution at the moment on Titan is Monte Carlo
% event generation and
detector simulation.
% Both types of task are
This type of task is mostly computational-intensive, requiring less than 1GB of
RAM at runtime and with small input data requirements. Detector simulation tasks
in ATLAS account for XX--XX\% of all jobs on WLCG, corresponding to X/X of the
current Grid capacity.\mtnote{Please provide numbers to replace `XX' and,
possibly, a diagram illustrating detector simulation stats compared to the whole
ATLAS computation activity.}

@@

% Event generation and
Detector simulation is part of the ATLAS production Monte Carlo (MC) workflow
(also known as MC production chain)~\cite{mcworkflow}. The MC workflow consists
of four main stages: event generation, detector simulation, digitization, and
reconstruction. Even generation creates sets of particle four-momenta via
different generators, e.g., PYTHIA~\cite{herwig} and HERWIG~\cite{herwig}. The
detector simulator is called Geant4~\cite{geant4} and simulates the interaction
of these particles with the sensitive material of the ATLAS detector. Each
interaction creates a so-called hit and all hits are collected and passed on for
digitalization where hits are further process to mimic the readout of the
detector. Finally, reconstruction operates local pattern recognition, creating
high-level objects like particles and jets.

@@

Compared to pilots, backfill has the disadvantage of limiting the amount of work
nodes that can be requested. Pilots are normal jobs: they can request as many
work nodes for as much time as a queue can offer. On the contrary, jobs sized on
the base of the backfill availability depend on the amount of work nodes that
cannot be given to any other job. Usually, backfill availability is a fraction
of the total capacity of the queue but the size of Titan mitigates this
limitation. Every year, about 10\% of Titan's capacity remains
unused~\cite{titan_utilization}, for a total of 30,000 cores, corresponding to
around 270M core hours per year, roughly 30\% of the overall capacity of WLCG.

@@

\mtnote{Please feel free to add details about backfill as needed.}
\mtnote{Please note: I know the following departs for the current name given to
the PanDA subsystem installed on Titan. I think there is a good reason to
call it a broker instead of a pilot, and I think I explained it in the previous
paragraphs. Please  take this just as a suggestion, something I would like to
discuss in our meeting.}

Given the communication requirements of PanDA Pilots and the unused capacity of
Titan, PanDA pilot was redesigned to serve as a job broker on Titan. Architected
as a stand-alone service to be executed on a DTN, this prototype called `PanDA
Broker' implements functionalities to: (i) interrogate Titan about backfill
availability; (ii) pull MC jobs and events; (iii) wrap jobs' payload into PBS
MPI jobs; (iv) submitting MPI PBS jobs to Titan's PBS batch system; and (v)
staging input/output files. Backfill querying, payload wrapping, and job
submission required a new implementation while job and events pulling, and file
staging were mostly inherited from PanDA Pilot.

>>>>

PanDA WMS has been designed to support distributed Grid
computing\sergeynote{specifically "distributed Grid computing"}\mtnote{done.}.
Executing ATLAS workloads or workflows involves concurrent and/or sequential
runs of possibly large amount of jobs, each requiring no or minimal
parallelization and no runtime communication. Thus, computing infrastructure
like WLCG have been designed to aggregate large amount of computing resources
across multiple sites.
%, often geographically dislocated.
\sergeynote{geographically distributed?}\mtnote{Better?} While each site may
deploy runtime message-passing capabilities, usually these are not used to
perform distributed computations.

@@

Currently, no HEP application can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks can be offloaded from
the Grid to Titan's large amount of cores. Further, when HEP tasks can be
partitioned to independent jobs, Titan work nodes can be used to execute up to
16 concurrent jobs' payload, one for each available core. Given these
constraints and challenges, the type of task most suitable for execution at the moment on Titan is Monte Carlo
% event generation and
detector simulation.
% Both types of task are
This type of task is mostly computational-intensive, requiring less than 2GB of
RAM at runtime and with small input data requirements. Detector simulation tasks
in ATLAS account for XX--XX\% of all jobs on WLCG, corresponding to X/X of the
current Grid capacity.\mtnote{Please provide numbers to replace `XX' and,
possibly, a diagram illustrating detector simulation stats compared to the whole
ATLAS computation activity.}

@@

% Event generation and
Detector simulation is part of the ATLAS production Monte Carlo (MC) workflow
(also known as MC production
chain)~\cite{rimoldi2006atlas,de2013delphes,ritsch2014atlas}. The MC workflow
consists of four main stages: event generation, detector simulation,
digitization, and reconstruction. Even generation creates sets of particle
four-momenta via different generators, e.g., PYTHIA~\cite{sjostrand2006pythia}
and HERWIG~\cite{corcella2001herwig}. The detector simulator is called
Geant4~\cite{agostinelli2003geant4} and simulates the interaction of these
particles with the sensitive material of the ATLAS detector. Each interaction
creates a so-called hit and all hits are collected and passed on for
digitalization where hits are further process to mimic the readout of the
detector. Finally, reconstruction operates local pattern recognition, creating
high-level objects like particles and jets.

@@

Compared to pilots, backfill has the disadvantage of limiting the amount of work
nodes that can be requested. Pilots are normal jobs: they can request as many
work nodes for as much time as a queue can offer. On the contrary, jobs sized on
the base of the backfill availability depend on the
% amount of work nodes that cannot be given to any other job.
any other job in the Titan's queue\sergeynote{any other job currently in the
Titan's queue?}\mtnote{Better?}. Usually, backfill availability is a fraction of
the total capacity of the queue but the size of Titan mitigates this limitation.
Every year, about 10\% of Titan's capacity remains
unused~\cite{titan_utilization}, for a total of 30,000 cores, corresponding to
around 270M core hours per year, roughly 30\% of the overall capacity of WLCG.

@@

\mtnote{Please feel free to add details about backfill as needed.}
\mtnote{Please note: I know the following departs for the current name given to
the PanDA subsystem installed on Titan. I think there is a good reason to call
it a broker instead of a pilot, and I think I explained it in the previous
paragraphs. Please  take this just as a suggestion, something I would like to
discuss in our meeting.} \sergeynote{It's not a pilot in conventional sense. I
call it an agent. It just happened that we were able to use full PanDA pilot's
code base to serve our purposes on Titan. that's why, by inertia, we still call
it a pilot.}\mtnote{Thank you. We would have a problem calling it `agent' as we
use that term to name the pilot of NGE. Would PanDA Broker work?}

Given the communication requirements of PanDA Pilots and the unused capacity of
Titan, PanDA pilot was redesigned to serve as a job broker on Titan. Architected
as a stand-alone service to be executed on a DTN, this prototype called `PanDA
Broker' implements functionalities to: (i) interrogate Titan about backfill
availability; (ii) pull MC jobs and events; (iii) wrap jobs' payload into PBS
MPI jobs; (iv) submitting MPI PBS jobs to Titan's PBS batch system and monitor
their execution;\sergeynote{Monitor job state and state transitions using
SAGA}\mtnote{Better?} and (v) staging input/output files. Backfill querying,
payload wrapping, and job submission required a new implementation while job and
events pulling, and file staging were mostly inherited from PanDA Pilot.

### 2017-03-07 PandaOverview.tex
First pass IV (former IV and V) + restructuring
Copy and address Sergey's comments + bibliography

The design of the Production and Distributed Analysis (PanDA) system started in
2005 to develop a WMS for the ATLAS experiment. PanDA went into production for
the LHC Run 1 between 2009 and 2013 and was then partially redesigned to be
deployed on Run 2, between 2015 and 2018. In this section, we summarize the
design, architecture, and execution process of the current state of the
practice.

@@

PanDA' data model uses events as the unit of data. Each event represents a
collision captured in the detector and one or more events can be packaged into
files or data source with other container abstractions. As with jobs, data have
both attributes and states, and some of the attributes are shared between events
and jobs so to create associations. Data are assumed to be stored centrally,
replicated when necessary and transferred over the network. This assumptions
applies to both input and output data.

@@

PanDA Server implementation consists of four components, all implemented in
Python and communicating via Python APIs: Task Buffer, Brokerage, Job
Dispatcher, and Data Service. Task Buffer is the management component of PanDA's
global queue and offers functionalities to: (i) store jobs into the global
queue; (ii) provide information about and update the state of each stored job.
The Brokerage component enables (i) matching job attributes with site and pilot
attributes; (ii) managing the dispatch of input data to processing sites; and
(iii) deriving PanDA's data pre-placement requirement.
% The job/pilot matching algorithm assigns a weight to each candidate pilot and
% after applying a set of policies, the best ten candidates are taken.
The Job Dispatcher implements functionalities to: (i) receive requests from
active pilots for new jobs to process; (ii) retrieving from Task Buffer the jobs
with highest priority that can be executed on the resource on which the
requesting pilots have been instantiated; (iii) creating a wrapper for the
selected job, tailored to the resource of the requesting pilot; and (iv)
monitoring the state of the job execution on the pilot, updating the Task Buffer
component. Finally, the Data Service component offers functionalities to: (i)
retrieve information from Task Buffer and Brokerage about the data requirements
of each job and the resource to which each job has been assigned; (ii) enable
interaction with the ATLAS Distributed Data Management (DDM)~\cite{atlas_ddm} to
make input files available to the jobs and manage their output data.

@@

\paragraph{\textbf{AutoPyFactory}} this subsystem implements a factory to submit
pilots locally and to both grid and cloud sites. Implemented to be executed as a
single demonized process, AutoPyFactory enables the concurrent spawning of
multiple job submission processes called APFQueue. Each APFQueue can communicate
with PanDA Server and the Brokerage module to collect information about the
state of the jobs it manages and, on the base of this information, submit new
pilots to one site per APFQueue. AutoPyFactory implements functionalities to
evaluate the number of pilots already instantiated on each site, and to apply a
consistent and adequate pressure on their batch queues. Submission is performed
via the Condor-G batch system~\cite{condorg}, enabling submission to multiple
type of sites. AutoPyFactory exposes a dedicated monitoring service, feeding
information to the PanDA Monitoring subsystem. A proxy manager is also run,
enabling certificate-based authentication and authorization.

@@

\paragraph{\textbf{PanDA Monitoring}} this subsystem is implemented as a
web-based, dashboard-style graphical application. It runs on Apache and
interacts with the back-end database of PanDA Server to enable persistence.
PanDA Monitoring enables users and site administrators to gather information
about the status of current and past jobs, data movement, and pilot factory.
PanDA Monitoring allows users access to all log files of each job, greatly
simplifying code debugging and failure analysis in a distributed computing
environment.

\paragraph{\textbf{JEDI}} this subsystem implements a workload and workflow
management for tasks instead of jobs, as defined by HEP users. Implemented as a
standalone service that interacts with Panda Server via its REST interface, JEDI
offers two main functionalities: partitioning tasks to jobs, and binding jobs to
resources via dedicated queues. JEDI partition tasks to jobs by matching tasks
requirements to resource capabilities. This matching is based on tasks
attributes, and static and dynamic information about the resources of each
available site. Dynamic information is acquired via a small number of `scout'
jobs that collect information about current resources and their performance.
Functionally, JEDI defines a set of queues associated to groups of resources
with specific capabilities. Tasks are assigned to groups of these queues, and
the jobs in which each task is partitioned are assigned to each queue, depending
on their requirements. JEDI replaces the job brokerage functionalities of the
Brokerage module of the PanDA Server but this module still handles communication
between the Task Buffer and the Data Service modules.
%, using it to enact the execution of the jobs defined by JEDI
% on the resources and sites decided by JEDI.

% -----------------------------------------------------------------------------
\subsection{Management and Execution}
\label{ssec:panda_exec}

@@

This approach introduced inefficiency both with usability and resource
utilization. Ideally, users should not have to reason in terms of jobs: Users
conceive analyses in terms of one or more, possibly related tasks; the `job'
abstraction is required exclusively by the execution middleware, i.e. PanDA.
Further, a static partitioning of tasks into jobs does not take into account the
heterogeneity and dynamicity of the resources of the pilots on which each job
will be executed.

@@

When PanDA Pilots become available on sites, they
% bootstrap, download their code, collect information about the execution
% environment and
call the Job Dispatcher module of the PanDA Server, requesting for a job to
execute \mtnote{Can a pilot request multiple jobs to execute concurrently?}. The
Job Dispatcher communicates with the Task Buffer, requesting for a job that is
bound to the site of that pilot and ready to be executed. Task Buffer checks the
global queue (i.e., the PanDA database) and if such a job is available, it
passes the job handler to the Job Dispatcher. The Job Dispatcher dispatches the
job to the PanDA Pilot.

@@

\mtnote{Think whether a comparison with other WMS based on the review we did
makes sense in this context.}

>>>>

The design of the Production and Distributed Analysis (PanDA) system started in
2005 to develop a WMS for the ATLAS experiment. PanDA went into production for
the LHC Run 1 on 2009 and was then partially redesigned to be deployed on Run 2,
on 2015. In this section, we summarize the design, architecture, and execution
process of the current state of the practice.

@@

PanDA' data model uses events as the unit of data. Each event represents a
collision captured in the detector and one or more events can be packaged into
files or data source with other container abstractions. As with jobs, data have
both attributes and states, and some of the attributes are shared between events
and jobs so to create associations. Raw, reconstruction, and simulation data are
assumed to be distributed across multiple and tiered sites and managed by the
ATLAS Distributed Data Management (DDM)~\cite{garonne2012atlas}. Datasets
required by each job are assumed to be replicated when necessary over the
network. This applies to both input and output data. \sergeynote{Raw data from
the ATLAS detector is captured at CERN and goes through reconstruction step to
produce data in secondary formats, smaller in volume and with more and more
physics content instead of raw detector hits. Consequently a second, custodial,
copy of raw data is being distributed in predetermined volumes to National T1
centers around the world (like BNL  T1 center). This insures low raw data
redundancy and safety. Multiple copies of chunks of secondary data sets (derived
from real raw data) are also distributed to ~100 centers around the word for
ease of access and improved availability for analysis. Then raw detector data is
stored on Tapes at CERN, just  in case. Most users, who do physics analysis,
rarely work with raw data, only a small group of experts needs to work with raw
data for specialized tasks like detector calibrations and such. There may be a
need to redo a reconstruction step, when errors in previous reconstruction
campaign are uncovered or improved detector understanding will lead for a better
quality physics. Then data stored on tape is recovered and re-processed again.
Simulation data is produced in many places anyway and is not centralized to
begin with. Here's the paper that describes current DDM state in ATLAS, data
replication strategies and such.
http://iopscience.iop.org/article/10.1088/1742-6596/396/3/032045/pdf DDM is a
complex and evolving topic in ATLAS.}\mtnote{Thank you. As we are describing the
design of PanDA I thought we should not enter into a too detailed description of
the DDM design. I eliminated the mistake I had made about `centralized data' but
left DDM just as a reference. Any better?}

@@

PanDA Server implementation consists of four components, all implemented in
Python and communicating via Python APIs: Task Buffer, Brokerage, Job
Dispatcher, and Data Service. Task Buffer is the management component of PanDA's
global queue and offers functionalities to: (i) store jobs into the global
queue; (ii) provide information about and update the state of each stored job.
The Brokerage component enables (i) matching job attributes with site and pilot
attributes; (ii) managing the dispatch of input data to processing sites; and
(iii) deriving PanDA's data pre-placement requirement.
% The job/pilot matching algorithm assigns a weight to each candidate pilot and
% after applying a set of policies, the best ten candidates are taken.
The Job Dispatcher implements functionalities to: (i) receive requests from
active pilots for new jobs to process; (ii) retrieving from Task Buffer the jobs
with highest priority that can be executed on the resource on which the
requesting pilots have been instantiated; (iii) creating a wrapper for the
selected job, tailored to the resource of the requesting pilot; and (iv)
monitoring the state of the job execution on the pilot, updating the Task Buffer
component. Finally, the Data Service component offers functionalities to: (i)
retrieve information from Task Buffer and Brokerage about the data requirements
of each job and the resource to which each job has been assigned; (ii) enable
interaction with the ATLAS DDM to make input files available to the jobs and
manage their output data.

@@

\paragraph{\textbf{AutoPyFactory}} this subsystem implements a factory to submit
pilots locally and to both grid and cloud
sites~\cite{caballero2012autopyfactory}. Implemented to be executed as a single
demonized process, AutoPyFactory enables the concurrent spawning of multiple job
submission processes called APFQueue. Each APFQueue can communicate with PanDA
Server and the Brokerage module to collect information about the state of the
jobs it manages and, on the base of this information, submit new pilots to one
site per APFQueue. AutoPyFactory implements functionalities to evaluate the
number of pilots already instantiated on each site, and to apply a consistent
and adequate pressure on their batch queues. Submission is performed via the
Condor-G batch system~\cite{frey2002condor}, enabling submission to multiple
type of sites. AutoPyFactory exposes a dedicated monitoring service, feeding
information to the PanDA Monitoring subsystem. A proxy manager is also run,
enabling certificate-based authentication and authorization.

@@

\paragraph{\textbf{PanDA Monitoring}} this subsystem is implemented as a
web-based, dashboard-style graphical application~\cite{klimentov2011atlas}. It
runs on Apache and interacts with the back-end database of PanDA Server to
enable persistence. PanDA Monitoring enables users and site administrators to
gather information about the status of current and past jobs, data movement, and
pilot factory. PanDA Monitoring allows users access to all log files of each
job, greatly simplifying code debugging and failure analysis in a distributed
computing environment.

\paragraph{\textbf{JEDI}} this subsystem implements a workload and workflow
management for tasks instead of jobs, as defined by HEP
users~\cite{borodin2015scaling}. Implemented as a standalone service that
interacts with Panda Server via its REST interface, JEDI offers two main
functionalities: partitioning tasks to jobs, and binding jobs to resources via
dedicated queues. JEDI partition tasks to jobs by matching tasks requirements to
resource capabilities. This matching is based on tasks attributes, and static
and dynamic information about the resources of each available site. Dynamic
information is acquired via a small number of `scout' jobs that collect
information about current resources and their performance. Functionally, JEDI
defines a set of queues associated to groups of resources with specific
capabilities. Tasks are assigned to groups of these queues, and the jobs in
which each task is partitioned are assigned to each queue, depending on their
requirements. JEDI replaces the job brokerage functionalities of the Brokerage
module of the PanDA Server but this module still handles communication between
the Task Buffer and the Data Service modules.
%, using it to enact the execution of the jobs defined by JEDI
% on the resources and sites decided by JEDI.

% -----------------------------------------------------------------------------
\subsection{Execution}
\label{ssec:panda_exec}

@@

This approach introduced inefficiency both with usability and resource
utilization~\cite{borodin2015unified}. Ideally, users should not have to reason
in terms of jobs: Users conceive analyses in terms of one or more, possibly
related tasks; the `job' abstraction is required exclusively by the execution
middleware, i.e. PanDA. Further, a static partitioning of tasks into jobs does
not take into account the heterogeneity and dynamicity of the resources of the
pilots on which each job will be executed.

@@

When PanDA Pilots become available on sites, they
% bootstrap, download their code, collect information about the execution
% environment and
call the Job Dispatcher module of the PanDA Server, requesting for a job to
execute \mtnote{Can a pilot request multiple jobs to execute
concurrently?}\sergeynote{In principle it should be able to do that, but that's not
how it is operated. One pilot - one job is a cleaner model in an environment
where resources are "owned" by the organization that runs pilots. Simplifies
many things. At least as far as I know.}. The Job Dispatcher communicates with
the Task Buffer, requesting for a job that is bound to the site of that pilot
and ready to be executed. Task Buffer checks the global queue (i.e., the PanDA
database) and if such a job is available, it passes the job handler to the Job
Dispatcher. The Job Dispatcher dispatches the job to the PanDA Pilot.

@@

\mtnote{Think whether a comparison with other WMS based on the review we did
makes sense in this context.}\sergeynote{You provided a nice description of the
PanDA system here.}
